Simplified and Verified: A Second Look at a Proof-Producing Union-Find Algorithm

### 2024-09-18 paper/CPP/paper.tex
Finish paper for submission

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,typedef,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive,record},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes,defines},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
\newcommand{\lstinlinew}[2][]{\lstinline[#1]{#2}}
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

% Acronyms
\usepackage[acronym]{glossaries}
\newacronym{adt}{ADT}{abstract data type}
\newacronym{uf}{UF}{Union-Find}
\newacronym{ufe}{UFE}{Union-Find-Explain}
\newacronym{lca}{LCA}{lowest common ancestor}
\newacronym{afp}{AFP}{Archive of Formal Proofs}
\newacronym{smt}{SMT}{Satisfiability Modulu Theories}

@@

\begin{abstract}
  TODO
\end{abstract}

@@

\section{Introduction}
The \acrfull{uf} data structure maintains the equivalence closure of a relation, where the relation is given as a sequence of pairs or, in terms of the \acrshort{uf} data structure, \opunion operations.
It is fundamental to efficiently implement well-known graph algorithms such as \citeauthor{mst}'s~\cite{mst} minimum spanning tree algorithm. 
There it tracks which vertices belong to the same connected component and are, in that sense, equivalent.
Beyond graph algorithms, its applicability extends to the domain of theorem proving as it routinely forms the basis of congruence closure algorithms, which are widely used by \acrfull{smt} solvers.
In order to increase their trustworthiness, current \acrshort{smt} solvers such as CVC5~\cite{cvc5}, E~\cite{eprover}, Vampire~\cite{vampire}, VeriT~\cite{verit}, and Z3~\cite{z3_proofs} are able to output detailed proofs if they determine their input formula to be unsatisfiable.

\begin{itemize}
  \item Based on \cite{congcl_proofs}
  \item Especially in the context of theorem proving certificates are useful
  \item Maybe cite order paper
\end{itemize}

\subsection{Contributions}
\begin{itemize}
  \item Formalisation of \acrshort{ufe} data structure
  \item Reduction of the complex recursion scheme to a simpler one that produces the same output
  \item Proof of soundness, completeness, and termination
  \item Abstract formalisation of union-find and union-find with size information hiding implementation details behind type definition
  \item Refinement of the \acrshort{ufe} to Imperative HOL to be able to export efficient code
\end{itemize}
\todo[inline]{We only show proof sketches, details in the formalisation}

\subsection{Related Work}
\begin{itemize}
  \item Union-Find as forest and union-by-size \cite{uf_by_size}
  \item Path compression \cite{uf_compress}
  \item Analysis of the run-time: $\mathcal{O}(n + m \alpha(m + n, n))$ for a sequence of at most $n - 1$ unions and $m$ finds \cite{uf_ub_improved}
  \item In Coq \cite{uf_coq} with time \cite{uf_coq_time}
  \item In Isabelle \cite{uf_isabelle} with time \cite{uf_isabelle} more recently viewed algebraically \cite{uf_isabelle_algebraic}.
\end{itemize}

@@

Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
The cardinality of a set \lstinline!A! is denoted by \lstinline!card A! and the image of \lstinline!A! under \lstinline!f! by \lstinline!f ` A!.

We use \lstinline!'a list! to describe the type of lists, which are constructed using the empty list \lstinline![]! constructor or the infix cons constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
\todo[inline]{Indexing operator}
\todo[inline]{Relation type synonym, Field of a relation}
\todo[inline]{Record notation}
\todo[inline]{Order on options}
\todo[inline]{Locale notation}
\todo[inline]{Lemma numbering}

We remark that $\longleftrightarrow$ is equivalent to \lstinline!$=$! on the type of Booleans \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
Meta-implication is denoted by \lstinline!$\Longrightarrow$! and a chain of implications
\begin{lstlisting}
  A$_\text{1}$ $\Longrightarrow$ $\cdots$ $\Longrightarrow$ A$_\text{k}$ $\Longrightarrow$ C
\end{lstlisting}
can be abbreviated by 
\begin{lstlisting}
  $\llbracket$ A$_\text{1}$;$\,\ldots\,$;A$_\text{k}$ $\rrbracket$ $\Longrightarrow$ C.
\end{lstlisting}

@@

We initialise the data structure by partitioning $A$ into singleton sets of elements,
so we have that $A = \{a_1\} \uplus \cdots \uplus \{a_n\}$.
Those sets are merged by subsequent \lstinline!union! operations\todo{use italics for union operation?} where \lstinline!union $a_i$ $a_j$! merges the set containing $a_i$ with the one that contains $a_j$.
Each set in the partition contains one particular element that serves as its representative.
We will denote the representative of an element $a$ in the \acrshort{uf} data structure $l$ as \lstinline!rep_of $l$ $a$!.
Accordingly, two elements have the same representative exactly when they belong to the same set in the partition.
For any element $a_i$, the \lstinline!find! operation just returns its representative \lstinline!rep_of $l$ $a_i$!.

@@

\subsection{In Isabelle/HOL\label{sec:uf_hol}}
The \acrshort{uf} algorithm is formalised in Isabelle/HOL by \citeauthor{uf_isabelle}~\cite{uf_isabelle}.
The code can be found in an entry~\cite{uf_isabelle_afp} of the \acrfull{afp}.\footnote{The code is in the theory file \texttt{Examples/Union\_Find.thy}.}
Later, \citeauthor{uf_isabelle}~\cite{uf_isabelle} also formalised the running time.\todo{Check citations}

The formalisation defines a function \lstinline|rep_of|, which, as described in the previous section, follows the parent pointers until we arrive the root, where the parent pointer is self-referential.
\begin{lstlisting}[columns=fullflexible]
function rep_of :: nat list $\Rightarrow$ nat $\Rightarrow$ nat where
rep_of l a =
  if l ! a = a then a else rep_of l (l ! a)
\end{lstlisting}
Looking closely at this definition, we see that this function is only well-defined for some inputs \lstinline|l| and \lstinline|a|:
for every element \lstinline|a < length l|, its parent must be in the list, i.e.\ we must have \lstinline|l ! a < length l|, and the parent pointers must be cycle-free in order for the function to terminate.
Functions in Isabelle/HOL must be total, so the \lstinline|function| command introduces a constant
\begin{lstlisting}
rep_of_dom :: (nat list $\times$ nat) $\Rightarrow$ bool  
\end{lstlisting}
that characterises the inputs for which \lstinline|rep_of| terminates.
Then, it adds \lstinline|rep_of_dom (l, a)| as a premise to the defining equation of \lstinline|rep_of|. 
The intuition above is cast into a predicate \lstinline|ufa_invar| that defines such well-formed lists \lstinline|l|.
\begin{lstlisting}
definition ufa_invar l = $\forall$a < length l.
  rep_of_dom (l, a) $\land$ l ! a < length l
\end{lstlisting}
Building on this formalisation, we encapsulate this \acrshort{uf} data structure into an \acrfull{adt} \lstinline|ufa| by defining
\begin{lstlisting}
typedef ufa = {uf. ufa_invar uf}
\end{lstlisting}
By lifting operations on the underlying list to the \acrshort{adt} using Isabelle's lifting infrastructure~\cite{lifting_transfer}, we obtain the operations
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ :: ufa $\Rightarrow$ nat rel!,
  \item \lstinline!ufa_rep_of :: ufa $\Rightarrow$ nat $\Rightarrow$ nat!, and
  \item \lstinline!ufa_union :: ufa $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ ufa!,
\end{itemize}
where \lstinline!nat rel = (nat $\times$ nat) set!.
The intended meaning of these operations is the following:
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ uf! is the partial equivalence relation represented by \lstinline!uf!,
  \item \lstinline!ufa_rep_of uf x! is the representative of the equivalence class that \lstinline!x! belongs to, and
  \item \lstinline!ufa_union uf x y! returns a \acrshort{uf} data structure where the equivalence classes of \lstinline!x! and \lstinline!y! are merged.
    This is implemented by updating the underlying list at index \lstinline!rep_of l x! to \lstinline|rep_of l y|.
\end{itemize}
More rigorously, the above operations fulfil the following properties:
\begin{lstlisting}
lemma ufa_rep_of uf x = ufa_rep_of uf y $\longleftrightarrow$
  (x, y) $\in$ ufa_$\alpha$ uf

@@

\section{Simple Certifying Union-Find Algorithm}
\begin{figure}
  \begin{gather*}
    \inferrule*[right=Assm]{\hbox{\lstinline|i < length us|} \\ \hbox{\lstinline|us ! i = (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ AssmP i : (x, y)|}} \\[0.5em]
    \inferrule*[right=Refl]{ }{\hbox{\lstinline|us $\ \vdash_=\ $ ReflP x : (x, x)|}} \\[0.5em]
    \inferrule*[right=Sym]{\hbox{\lstinline|us $\ \vdash_=\ $ p : (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ SymP p : (y, x)|}} \\[0.5em]
    \inferrule*[right=Trans]{\hbox{\lstinline|us $\ \vdash_=\ $ p1 : (x, y)|} \\ \hbox{\lstinline|us $\ \vdash_=\ $ p2 (y, z)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ p1 $\ \TransP\ $ p2 : (x, z)|}}
  \end{gather*}
  \caption{The data type \lstinline|eq_prf| of certificates and its corresponding system of inference rules $\vdash_=$.\label{fig:eq_prf}}
\end{figure}
\todo[inline]{Use equivalence instead of equality, Use equivalence operator instead of pairs, Maybe use cong symbol}
Building on the \acrshort{uf} \acrshort{adt} from the previous section, we now develop a simple \lstinline|explain| operation that,
for a given list of equations \lstinline|us|, takes two elements \lstinline|x| and \lstinline|y| and produces a certificate that \lstinline|x = y| modulo \lstinline|us|.
The certificate is given in terms of a data type \lstinline|eq_prf| with its corresponding system $\vdash_=$ of inference rules as seen in \autoref{fig:eq_prf}.
As expected, we have inference rules that utilise the reflexivity, symmetry, and transitivity of equality as well as an assumption rule.
To improve readability, we use the infix operator $\bigtriangledown$ to denote the proof term for transitivity.

@@

We prove that $\vdash_=$ is sound and complete with respect to the equivalence relation induced by \lstinline|us|, i.e.\ the equivalence closure of \lstinline|us|.
In Isabelle, we define\todo{Introduce inverse and rtrancl?}
\begin{lstlisting}
definition symcl r = r $\cup$ r$^{-1}$
definition equivcl r = (symcl r)$^*$
\end{lstlisting}
and prove
\begin{lstlisting}
lemma $\vdash_=$_sound:
assumes as $\vdash_=$ p : (x, y)
shows (x, y) $\in$ equivcl (set as)

@@

\begin{figure*}
  \centering
  \begin{lstlisting}
function explain :: (nat $\times$ nat) list $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ nat eq_prf where
  explain [] x _ = ReflP x
| explain (us @ [(a, b)]) x y =
    let uf = ufa_unions uf_init us; a_b_P = AssmP (length us)
    in if ufa_rep_of uf x = ufa_rep_of uf y then explain us x y
      else if ufa_rep_of uf x = ufa_rep_of uf a then explain us x a $\TransP$ a_b_P $\TransP$ explain us y b
      else explain us x b $\TransP$ SymP a_b_P $\TransP$ explain us a y

definition explain_partial us x y =
  if (x, y) $\in$ equivcl (set us) then Some (explain us x y) else None
  \end{lstlisting}
  \caption{A simple implementation of the \lstinline|explain| operation. It operates on an arbitrary initial \acrshort{uf} data structure \lstinline|uf_init|.\label{fig:explain}}
\end{figure*}
\todo[inline]{termination of \lstinline|explain|}
Although \lstinline|explain| takes a list of equations \lstinline|us|, its implementation actually works with an instance \lstinline|uf| of the \acrshort{uf} \acrshort{adt} that results from applying \lstinline|us| as unions to an initial \acrshort{uf} data structure \lstinline|uf_init|.
Accordingly, we define
\begin{lstlisting}
definition ufa_unions =
  foldl ($\lambda$uf (x, y). ufa_union uf x y)
\end{lstlisting}
so we have \lstinline|uf = ufa_unions uf_init us| in terms of the constants above.

@@

Otherwise, the list of unions must be equal to \lstinline|us @ [(a, b)]| for some \lstinline|us|, \lstinline|a|, and \lstinline|b|.
We distinguish between two cases:
\begin{itemize}
  \item The elements \lstinline|x| and \lstinline|y| are already equal modulo \lstinline|us|, so we proceed recursively with \lstinline|us|.
  \item In the case that the equation \lstinline|a = b| is necessary for \lstinline|x = y| to hold, we either have \lstinline|x = a| and \lstinline|b = y| or \lstinline|x = b| and \lstinline|a = y| modulo \lstinline|us|.
    Assuming that the former holds --- the other case is symmetric --- we recursively construct the certificates for \lstinline|x = a| and \lstinline|b = y|.
    Together with the assumption that \lstinline|a = b|, we obtain \lstinline|x = y| by transitivity.
\end{itemize}

@@

The \lstinline|explain| function is not sound, though.
This is because it always returns a certificate, even if \lstinline|x| and \lstinline|y| are not equal modulo \lstinline|us|.
To account for this case, we wrap \lstinline|explain| into a partial function \lstinline|explain_partial| (cf.\ \autoref{fig:explain}) that fails if \lstinline|x = y| is not provable.
Soundness and completeness can then be lifted from the soundness of $\vdash_=$ and the completeness of \lstinline|explain|, respectively.
At first glance, it might seem vacuous to define \lstinline|explain_partial| using \lstinline|equivcl|;
however, \lstinline|equivcl| can actually be implemented using \acrshort{uf} operations as the following lemma demonstrates.
\begin{lstlisting}
lemma
defines uf = ufa_unions uf_init us
shows (x, y) $\in$ equivcl (set us) $\leftrightarrow$ x = y $\lor$
  x $\in$ Field (ufa_$\alpha$ uf) $\land$ y $\in$ Field (ufa_$\alpha$ uf) $\land$
  ufa_rep_of uf x = ufa_rep_of uf y
\end{lstlisting}
\todo[inline]{Improve explanation above? Segue?}

>>>>

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{context,type_synonym,typedef,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive,record},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes,defines,begin,end},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
\newcommand{\lstinlinew}[2][]{\lstinline[#1]{#2}}
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

% Acronyms
\usepackage[acronym]{glossaries}
\newacronym{adt}{ADT}{abstract data type}
\newacronym{uf}{UF}{Union-Find}
\newacronym{ufe}{UFE}{Union-Find-Explain}
\newacronym{lca}{LCA}{lowest common ancestor}
\newacronym{afp}{AFP}{Archive of Formal Proofs}
\newacronym{smt}{SMT}{Satisfiability Modulo Theories}

@@

\begin{abstract}
Using Isabelle/HOL, we verify an Union-Find data structure with an explain operation due to \citeauthor{congcl_proofs}.
We devise a simpler, more naive version of the explain operation whose soundness and completeness is easy to verify.
Then, we prove the original formulation of the explain operation to be equal to our version.
Finally, we refine this data structure to Imperative HOL, enabling us to export efficient imperative code.
The formalisation provides a stepping stone towards the verification of proof-producing congruence closure algorithms which are a core ingredient of \acrfull{smt} solvers.
\end{abstract}

@@

\section{Introduction}
The \acrfull{uf} data structure maintains the equivalence closure of a relation, where the relation is given as a sequence of pairs or, in terms of the \acrshort{uf} data structure, \opunion{} operations.
It is fundamental to efficiently implement well-known graph algorithms such as \citeauthor{mst}'s~\cite{mst} minimum spanning tree algorithm. 
There it tracks which vertices belong to the same connected component and are, in that sense, equivalent.
Beyond graph algorithms, its applicability extends to the domain of theorem proving as it routinely forms the basis of congruence closure algorithms, which are widely used by \acrfull{smt} solvers.
In order to increase their trustworthiness, current \acrshort{smt} solvers such as CVC5~\cite{cvc5}, E~\cite{eprover}, Vampire~\cite{vampire}, VeriT~\cite{verit}, and Z3~\cite{z3_proofs} are able to output detailed proofs when they determine their input formula to be unsatisfiable.
To produce these proofs, it is crucial to have congruence closure algorithms that can efficiently explain why they consider two terms to be equal.
The first such algorithm was presented by \citeauthor{congcl_proofs}~\cite{congcl_proofs,congcl_fast_extensions}, 
who also extend the \acrshort{uf} data structure with an \opexplain{} operation to obtain an \acrfull{ufe} data structure as part of their work.
Verifying this data structure is the focus of our paper.

\subsection{Contributions}
The contributions of this paper are the following:
\begin{itemize}
  \item We present, to our knowledge, the first formalisation of the \acrshort{ufe} data structure as introduced by \citeauthor{congcl_proofs}.
    In their work, they present two variants of this data structure.
    Here, we only formalise the first variant, leaving the other for future work.
  \item We devise a simpler, more naive version of the \opexplain{} operation, for which soundness and completeness is easier to prove. 
  \item Then, we prove the original version of the \opexplain{} operation to be extensionally equal to the simple one.
  \item Based on an existing formalisation of the \acrshort{uf} data structure by \citeauthor{uf_isabelle}~\cite{uf_isabelle},
    we develop a more abstract formalisation of the data structure hiding implementation details. 
  \item Finally, we refine the \acrshort{ufe} data structure to Imperative HOL~\cite{imperative_hol} using \citeauthor{uf_isabelle}'s~\cite{uf_isabelle} separation logic framework.
    This enables us to export efficient imperative code.
\end{itemize}
The formalisation is available as supplementary material to this paper.
As everything is verified, we largely skip over the proofs in this paper and only sometimes provide proof sketches in order to illustrate the structure of the formalisation.

\subsection{Related Work}
Efficient implementations of the \acrshort{uf} data structure have been known for a long time.
In particular, \citeauthor{uf_by_size} represent the data structure as a forest of rooted trees and propose the union-by-size heuristic, which gives $\mathcal{O}(\log n)$ running time for \opunion{} and \opfind{} for a data struture over $n$ elements.
Another heuristic, called path compression, was presented by \citeauthor{uf_compress}~\cite{uf_compress}.
Analysing the complexity of the data structure when combining both heuristics turned to be challenging but \citeauthor{uf_ub}~\cite{uf_ub} and \citeauthor{uf_ub_improved}~\cite{uf_ub_improved} eventually proved an amortised running time of $\mathcal{O}(n + m\, \alpha(m + n, n))$ for a sequence of at most $n - 1$ unions and $m$ finds where $\alpha$ is the inverse Ackermann function. 
This means that any one operation runs in almost constant time, amortised.

The \acrshort{ufe} data structure~\cite{congcl_proofs} itself has received limited attention in the literature.
It does, however, form the basis of proof-producing congruence closure algorithms, which are crucial in the field of \acrshort{smt} solving.
There, they remain an active area of research;
for example, when we are interested in efficiently finding small proofs~\cite{congcl_small_proofs}.

In the context of interactive theorem proving, there is a formalisation of the \acrshort{uf} data structure in Coq~\cite{uf_coq}.
Its amortised complexity is analysed by \citeauthor{uf_coq_time}~\cite{uf_coq_time} in a separation logic with time credits.
Similarly, in Isabelle, there is a formalisation of the data structure~\cite{uf_isabelle} that was later extended with a complexity analysis by \citeauthor{uf_isabelle_time}~\cite{uf_isabelle_time}.
More recently, there is formalisation by \citeauthor{uf_isabelle_algebraic}~\cite{uf_isabelle_algebraic} taking a relation-algebraic view.

@@

Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
The cardinality of a set \lstinline!A! is denoted by \lstinline!card A!.

We use \lstinline!'a list! to describe the type of lists, which are constructed using the empty list \lstinline![]! constructor or the infix cons constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
We write \lstinline|xs ! i| to access the \lstinline|i|-th element of the list \lstinline|xs|.

To represent partial values of type \lstinline|'a|, we use the type \lstinline|'a option| with the constructors \lstinline|None| and \lstinline|Some|.
We also define an order on this type by letting \lstinline|None| be smaller than \lstinline|Some a| and lifting the order on the underlying type \lstinline|'a|, i.e.\ we have that
\begin{lstlisting}
Some x $\leq$ Some y $\longleftrightarrow$ x $\leq$ y.
\end{lstlisting}

Relations are denoted with the type synonym \lstinline|'a rel|, which expands to \lstinline|('a $\times$ 'a) set|.
For a relation \lstinline|r|, \lstinline|Field r| are those elements that occur as a component of a pair \lstinline|p $\in$ r|.
Furthermore, we use \lstinline|r$^{-1}$| to denote the inverse and \lstinline|r$^*$| to denote the reflexive transitive closure of \lstinline|r|.

We remark that $\longleftrightarrow$ is equivalent to \lstinline!$=$! on the type of Booleans \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
%\todo[inline]{Record notation}
%\todo[inline]{Locale notation}
%\todo[inline]{Lemma numbering}
%Meta-implication is denoted by \lstinline!$\Longrightarrow$! and a chain of implications
%\begin{lstlisting}
%  A$_\text{1}$ $\Longrightarrow$ $\cdots$ $\Longrightarrow$ A$_\text{k}$ $\Longrightarrow$ C
%\end{lstlisting}
%can be abbreviated by 
%\begin{lstlisting}
%  $\llbracket$ A$_\text{1}$;$\,\ldots\,$;A$_\text{k}$ $\rrbracket$ $\Longrightarrow$ C.
%\end{lstlisting}

@@

We initialise the data structure by partitioning $A$ into singleton sets of elements,
so we have that $A = \{a_1\} \uplus \cdots \uplus \{a_n\}$.
Those sets are merged by subsequent \opunion{} operations where \lstinline!union $a_i$ $a_j$! merges the set containing $a_i$ with the one that contains $a_j$.
Each set in the partition contains one particular element that serves as its representative.
We will denote the representative of an element $a$ in the \acrshort{uf} data structure $l$ as \lstinline!rep_of $l$ $a$!.
Accordingly, two elements have the same representative exactly when they belong to the same set in the partition.
For any element $a_i$, the \lstinline!find! operation just returns its representative \lstinline!rep_of $l$ $a_i$!.

@@

\subsection{In Isabelle/HOL\label{sec:uf_hol}}
The \acrshort{uf} algorithm is formalised in Isabelle/HOL by \citeauthor{uf_isabelle}~\cite{uf_isabelle}.
The code can be found in an entry~\cite{uf_isabelle_afp} of the \acrfull{afp}.\footnote{The code is in the theory file \texttt{Examples/Union\_Find.thy}.}

The formalisation defines a function \lstinline|rep_of|, which, as described in the previous section, follows the parent pointers until we arrive the root, where the parent pointer is self-referential.
\begin{lstlisting}[columns=fullflexible]
function rep_of :: nat list $\Rightarrow$ nat $\Rightarrow$ nat where
rep_of l a =
  if l ! a = a then a else rep_of l (l ! a)
\end{lstlisting}
Looking closely at this definition, we see that this function is only well-defined for some inputs \lstinline|l| and \lstinline|a|:
for every element \lstinline|a < length l|, its parent must be in the list, i.e.\ we must have \lstinline|l ! a < length l|, and the parent pointers must be cycle-free in order for the function to terminate.
Functions in Isabelle/HOL must be total, so the \lstinline|function| command introduces a constant
\begin{lstlisting}
rep_of_dom :: (nat list $\times$ nat) $\Rightarrow$ bool  
\end{lstlisting}
that characterises the inputs for which \lstinline|rep_of| terminates.
Then, it adds \lstinline|rep_of_dom (l, a)| as a premise to the defining equation of \lstinline|rep_of|. 
The intuition above is cast into a predicate \lstinline|ufa_invar| that defines such well-formed lists \lstinline|l|.
\begin{lstlisting}
definition ufa_invar l = $\forall$a < length l.
  rep_of_dom (l, a) $\land$ l ! a < length l
\end{lstlisting}
Building on this formalisation, we encapsulate this \acrshort{uf} data structure into an \acrfull{adt} \lstinline|ufa| by defining
\begin{lstlisting}
typedef ufa = {uf. ufa_invar uf}
\end{lstlisting}
By lifting operations on the underlying list to the \acrshort{adt} using Isabelle's lifting infrastructure~\cite{lifting_transfer}, we obtain the operations
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ :: ufa $\Rightarrow$ nat rel!,
  \item \lstinline!ufa_rep_of :: ufa $\Rightarrow$ nat $\Rightarrow$ nat!, and
  \item \lstinline|ufa_init :: nat $\Rightarrow$ ufa|,
  \item \lstinline!ufa_union :: ufa $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ ufa!,
\end{itemize}
where \lstinline!nat rel = (nat $\times$ nat) set!.
The intended meaning of these operations is the following:
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ uf! is the partial equivalence relation represented by \lstinline!uf!,
  \item \lstinline!ufa_rep_of uf x! is the representative of the equivalence class that \lstinline!x! belongs to,
  \item \lstinline|ufa_init n| initialises the data structure with \lstinline|n| elements with each element being its own representative, and
  \item \lstinline!ufa_union uf x y! returns a \acrshort{uf} data structure where the equivalence classes of \lstinline!x! and \lstinline!y! are merged.
    This is implemented by updating the underlying list at index \lstinline!rep_of l x! to \lstinline|rep_of l y|.
\end{itemize}
More rigorously, the above operations fulfil the following properties:
\begin{lstlisting}
lemma ufa_rep_of uf x = ufa_rep_of uf y $\longleftrightarrow$
  (x, y) $\in$ ufa_$\alpha$ uf

@@

\section{Simple Certifying Union-Find Algorithm\label{sec:ufe_simple}}
\begin{figure}
  \begin{gather*}
    \inferrule*[right=Assm]{\hbox{\lstinline|i < length us|} \\ \hbox{\lstinline|us ! i = (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ AssmP i : (x, y)|}} \\[0.5em]
    \inferrule*[right=Refl]{ }{\hbox{\lstinline|us $\ \vdash_=\ $ ReflP x : (x, x)|}} \\[0.5em]
    \inferrule*[right=Sym]{\hbox{\lstinline|us $\ \vdash_=\ $ p : (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ SymP p : (y, x)|}} \\[0.5em]
    \inferrule*[right=Trans]{\hbox{\lstinline|us $\ \vdash_=\ $ p1 : (x, y)|} \\ \hbox{\lstinline|us $\ \vdash_=\ $ p2 (y, z)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ p1 $\ \TransP\ $ p2 : (x, z)|}}
  \end{gather*}
  \caption{The data type \lstinline|eq_prf| of certificates and its corresponding system of inference rules $\vdash_=$.\label{fig:eq_prf}}
\end{figure}
Building on the \acrshort{uf} \acrshort{adt} from the previous section, we now develop a simple \lstinline|explain| operation that,
for a given list of equations \lstinline|us|, takes two elements \lstinline|x| and \lstinline|y| and produces a certificate that \lstinline|x = y| modulo \lstinline|us|.
The certificate is given in terms of a data type \lstinline|eq_prf| with its corresponding system $\vdash_=$ of inference rules as seen in \autoref{fig:eq_prf}.
As expected, we have inference rules that utilise the reflexivity, symmetry, and transitivity of equality as well as an assumption rule.
To improve readability, we use the infix operator $\bigtriangledown$ to denote the proof term for transitivity.

@@

We prove that $\vdash_=$ is sound and complete with respect to the equivalence relation induced by \lstinline|us|, i.e.\ the equivalence closure of \lstinline|us|.
In Isabelle, we define
\begin{lstlisting}
definition symcl r = r $\cup$ r$^{-1}$
definition equivcl r = (symcl r)$^*$
\end{lstlisting}
and prove
\begin{lstlisting}
lemma $\vdash_=$_sound:
assumes as $\vdash_=$ p : (x, y)
shows (x, y) $\in$ equivcl (set as)

@@

\begin{figure*}
  \centering
  \begin{lstlisting}
context fixes uf_init :: ufa
begin

function explain :: (nat $\times$ nat) list $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ nat eq_prf where
  explain [] x _ = ReflP x
| explain (us @ [(a, b)]) x y =
    let uf = ufa_unions uf_init us; a_b_P = AssmP (length us)
    in if ufa_rep_of uf x = ufa_rep_of uf y then explain us x y
       else if ufa_rep_of uf x = ufa_rep_of uf a then explain us x a $\TransP$ a_b_P $\TransP$ explain us y b
       else explain us x b $\TransP$ SymP a_b_P $\TransP$ explain us a y

definition explain_partial us x y =
  if (x, y) $\in$ equivcl (set us) then Some (explain us x y) else None

end
  \end{lstlisting}
  \caption{A simple implementation of the \lstinline|explain| operation. We fix an arbitrary initial \acrshort{uf} data structure \lstinline|uf_init| that the operation works on.\label{fig:explain}}
\end{figure*}
Although \lstinline|explain| takes a list of equations \lstinline|us|, its implementation actually works with an instance \lstinline|uf| of the \acrshort{uf} \acrshort{adt} that results from applying \lstinline|us| as unions to an initial \acrshort{uf} data structure \lstinline|uf_init|.
Accordingly, we define
\begin{lstlisting}
definition ufa_unions =
  foldl ($\lambda$uf (x, y). ufa_union uf x y)
\end{lstlisting}
so we have \lstinline|uf = ufa_unions uf_init us| in terms of the constants above.

@@

Otherwise, the list of unions is equal to \lstinline|us @ [(a, b)]| for some \lstinline|us|, \lstinline|a|, and \lstinline|b|.
We distinguish between two cases:
\begin{itemize}
  \item The elements \lstinline|x| and \lstinline|y| are already equal modulo \lstinline|us|, so we proceed recursively with \lstinline|us|.
  \item In the case that the equation \lstinline|a = b| is necessary for \lstinline|x = y| to hold, we either have \lstinline|x = a| and \lstinline|b = y| or \lstinline|x = b| and \lstinline|a = y| modulo \lstinline|us|.
    Assuming the former holds --- the other case is symmetric --- we recursively construct the certificates for \lstinline|x = a| and \lstinline|b = y|.
    Together with the assumption that \lstinline|a = b|, we obtain \lstinline|x = y| by transitivity.
\end{itemize}
The termination of \lstinline|explain| is easily proven as the length of the list of unions decreases in each call.

@@

The \lstinline|explain| function is not sound, though.
This is because it always returns a certificate, even if \lstinline|x| and \lstinline|y| are not equal modulo \lstinline|us|.
To account for this case, we wrap \lstinline|explain| into a partial function \lstinline|explain_partial| (cf.\ \autoref{fig:explain}) that fails if \lstinline|x = y| is not provable.
Soundness and completeness can then be lifted from the soundness of $\vdash_=$ and the completeness of \lstinline|explain|, respectively.
At first glance, it might seem vacuous to define \lstinline|explain_partial| using \lstinline|equivcl|;
however, \lstinline|equivcl| can actually be implemented using \acrshort{uf} operations as the following lemma demonstrates.
\begin{lstlisting}
lemma
defines uf = ufa_unions uf_init us
shows (x, y) $\in$ equivcl (set us) $\leftrightarrow$ x = y $\lor$
  x $\in$ Field (ufa_$\alpha$ uf) $\land$ y $\in$ Field (ufa_$\alpha$ uf) $\land$
  ufa_rep_of uf x = ufa_rep_of uf y
\end{lstlisting}
